---
title: "OA_Multi_Omics_Rmarkdown_Cytokines_and_SCFA"
author: "Weilan Wang 2026@ UofC"
date: "2026-06-11"
output: html_document
---


```{r library setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd('C:/Users/Weilan Wang/Desktop/weilan_R')
## install.packages("viridis")
## install.packages("multiblock")
## install.packages("lmerTest")
## install.packages("emmeans")
## install.packages("Matrix", type = "source")
## BiocManager::install("DESeq2")
## BiocManager::install("EnhancedVolcano")
## BiocManager::install("MOFA2")
## BiocManager::install("limma")
## BiocManager::install("phyloseq")
## BiocManager::install("microViz")
## BiocManager::install("MOFAdata")
## BiocManager::install(c("phyloseq", "microbiome", "ComplexHeatmap"), update = FALSE)
## install.packages("microViz",repos = c(davidbarnett = "https://david-barnett.r-universe.dev", getOption("repos")))

library(broom.mixed)
library(car)
library(caret)
library(circlize)
library(ComplexHeatmap)
library(cowplot)
library(data.table)
library(DESeq2)
library(doParallel)
library(dplyr)
library(egg)
library(EnhancedVolcano)
library(emmeans)
library(GGally)
library(ggcorrplot)
library(ggraph)
library(ggnewscale)
library(ggforce)
library(ggplot2)
library(ggpubr)
library(gridExtra)
library(grid)
library(git2r)
library(Hmisc)
library(igraph)
library(lattice)
library(lme4)
library(lmerTest)
library(limma)
library(lsmeans)
library(magrittr)
library(Matrix)
library(mgcv)
library(microViz)
library(MOFA2)
library(MOFAdata)
library(multiblock)
library(MuMIn)
library(nlme)
library(pheatmap)
library(phyloseq)
library(purrr)
library(RColorBrewer)
library(readr)
library(readxl)
library(reshape2)
library(rstatix)
library(tidyr)
library(tidyverse)
library(vegan)
library(viridis)

```

## R Markdown

```{r theme setup for figures, echo=FALSE}
mytheme1 <- theme(panel.background = element_rect(fill = "white", color="black"), 
                  axis.title = element_text(face="bold",size=12, color = "black"),
                  axis.text.y  = element_text(face="bold",size=12, color = "black"),
                  axis.text.x  = element_text(face="bold",size=12, color = "black"),
                  panel.grid.minor.y=element_blank(), 
                  panel.grid.minor.x=element_blank(),
                  panel.grid.major.y=element_blank(),
                  panel.grid.major.x=element_blank(),
                  strip.background = element_blank(),
                  strip.text.x = element_text(face="bold",size=14, color = "black"),
                  legend.position = "top",
                  legend.box = "horizontal",
                  legend.title= element_blank(),
                  legend.text =element_text(size=14),
                  legend.key = element_blank())

mytheme2 <- theme(panel.background = element_rect(fill = "white", color = NA), 
                  panel.border = element_blank(), 
                  axis.line.x = element_line(color = "black"),  # Keep the x-axis line
                  axis.line.y = element_line(color = "black"),
                  axis.title = element_text(face="bold",size=12, color = "black"),
                  axis.text.y  = element_text(face="bold",size=12, color = "black"),
                  axis.text.x  = element_text(face="bold",size=12, color = "black"),
                  panel.grid.minor.y=element_blank(), 
                  panel.grid.minor.x=element_blank(),
                  panel.grid.major.y=element_blank(),
                  panel.grid.major.x=element_blank(),
                  strip.background = element_blank(),
                  strip.text.x = element_blank(),
                  strip.text.y = element_blank(),
                  legend.position = "top",
                  legend.box = "vertical",
                  legend.title= element_text(face="bold",size=10, color = "black"),
                  legend.text =element_text(size=10),
                  legend.key = element_blank())

mytheme3 <- theme(panel.background = element_rect(fill = "white", color = NA), 
                  panel.border = element_blank(), 
                  axis.line.x = element_line(color = "black"),  # Keep the x-axis line
                  axis.line.y = element_line(color = "black"),
                  axis.title = element_text(face="bold",size=12, color = "black"),
                  axis.text.y  = element_text(size=10, color = "black"),
                  axis.text.x  = element_text(size=10, color = "black"),
                  panel.grid.minor.y=element_blank(), 
                  panel.grid.minor.x=element_blank(),
                  panel.grid.major.y=element_blank(),
                  panel.grid.major.x=element_blank(),
                  strip.background = element_blank(),
                  strip.text.x = element_blank(),
                  strip.text.y = element_blank(),
                  legend.position = "none",
                  legend.box = "vertical",
                  legend.title= element_text(face="bold",size=10, color = "black"),
                  legend.text =element_text(size=10),
                  legend.key = element_blank())

mytheme4 <- theme(panel.background = element_rect(fill = "white", color = NA), 
                  panel.border = element_rect(color = "black", fill = NA),  # Use element_rect for panel.border
                  axis.line.x = element_line(color = "black"),  # Keep the x-axis line
                  axis.line.y = element_line(color = "black"),
                  axis.title = element_text(face="bold", size=12, color = "black"),
                  axis.text.y  = element_text(size=10, color = "black"),
                  axis.text.x  = element_text(size=10, color = "black"),
                  panel.grid.minor.y = element_blank(), 
                  panel.grid.minor.x = element_blank(),
                  panel.grid.major.y = element_line(color = "grey", linetype = "dashed"),
                  panel.grid.major.x = element_line(color = "grey", linetype = "dashed"),
                  strip.background = element_rect(),
                  strip.text.x = element_text(size=10, color = "black"),
                  strip.text.y = element_blank(),
                  legend.position = "right",
                  legend.box = "vertical",
                  legend.title = element_text(size=10, color = "black"),
                  legend.text = element_text(size=10),
                  legend.key = element_rect())
```

```{r Serum Cytokine, LPS & Fecal SCFA Stats & Figures}
Raf_OA_phenotype_data <- read.csv("Raf_OA_phenotypes.csv")
threshold <- 3 

# Filter out outliers for each numeric variable within each Group
Raf_OA_phenotype_data_outlier_filtered <- Raf_OA_phenotype_data %>%
  group_by(Group) %>%
  mutate(across(where(is.numeric), ~ ifelse(abs(. - mean(., na.rm = TRUE)) <= threshold * sd(., na.rm = TRUE), ., NA))) %>%
  ungroup()

str(Raf_OA_phenotype_data_outlier_filtered)

Raf_OA_phenotype_data_outlier_filtered$Group <- factor(Raf_OA_phenotype_data_outlier_filtered$Group, 
                                            levels = c('Placebo_M0', 'Prebiotic_M0',
                                                       'Placebo_M3', 'Prebiotic_M3', 
                                                       'Placebo_M6', 'Prebiotic_M6'))

Raf_OA_phenotype_participant_counts <- table(Raf_OA_phenotype_data_outlier_filtered$Participant)
Raf_OA_phenotype_data_outlier_filtered$Diets <- factor(Raf_OA_phenotype_data_outlier_filtered$Diets)
Raf_OA_phenotype_data_outlier_filtered$Months <- factor(Raf_OA_phenotype_data_outlier_filtered$Months)

Raf_OA_phenotypes <- names(Raf_OA_phenotype_data_outlier_filtered)[
  which(names(Raf_OA_phenotype_data_outlier_filtered) == "TotalMass"):ncol(Raf_OA_phenotype_data_outlier_filtered)]

all_results <- data.frame()

for (phenotype in Raf_OA_phenotypes) {
  formula <- as.formula(paste(phenotype, "~ Diets * Months + (1 | Participant)"))
  lme_model <- lmer(formula, data = Raf_OA_phenotype_data_outlier_filtered, REML = FALSE)
  emms <- emmeans(lme_model, ~ Diets * Months)
  pairwise_comparisons <- contrast(emms, method = "pairwise", simple = "each")
  pairwise_comparisons_adjusted <- summary(pairwise_comparisons, adjust = "bonferroni")
  phenotype_results <- as.data.frame(pairwise_comparisons_adjusted)
  phenotype_results$Phenotype <- phenotype
  all_results <- bind_rows(all_results, phenotype_results)
  print(paste("Processed results for", phenotype))
}

write.csv(all_results, "Raf_OA_Raw_Phenotypes_Pairwise_Comparisons.csv", row.names = FALSE)
write.csv(Raf_OA_phenotype_data_outlier_filtered, 
          "Raf_OA_phenotype_data_outlier_filtered.csv", row.names = FALSE)

#### phenotype figures- raw ####
Raf_OA_phenotype_data_long <- Raf_OA_phenotype_data_outlier_filtered %>%
  pivot_longer(
    cols = TotalMass:Knee_pain, 
    names_to = "Params_catlog",
    values_to = "Params_Value"
  ) %>%
  mutate(Params_type = case_when(
    Params_catlog %in% names(Raf_OA_phenotype_data_outlier_filtered)
    [which(names(Raf_OA_phenotype_data_outlier_filtered) == "TotalMass"):which(names(Raf_OA_phenotype_data_outlier_filtered) == "TrunkFat.")] ~ "Body Composition",
    Params_catlog %in% names(Raf_OA_phenotype_data_outlier_filtered)
    [which(names(Raf_OA_phenotype_data_outlier_filtered) == "GM_CSF"):which(names(Raf_OA_phenotype_data_outlier_filtered) == "IL_10")] ~ "Serum Proinflammatory Markers",
    Params_catlog %in% names(Raf_OA_phenotype_data_outlier_filtered)
    [which(names(Raf_OA_phenotype_data_outlier_filtered) == "Acetate"):which(names(Raf_OA_phenotype_data_outlier_filtered) == "Valerate")] ~ "SCFA",
    Params_catlog %in% names(Raf_OA_phenotype_data_outlier_filtered)
    [which(names(Raf_OA_phenotype_data_outlier_filtered) == "SitToStand"):which(names(Raf_OA_phenotype_data_outlier_filtered) == "TimeUpGo")] ~ "Function Test",
    Params_catlog %in% names(Raf_OA_phenotype_data_outlier_filtered)
    [which(names(Raf_OA_phenotype_data_outlier_filtered) == "LPS")] ~ "LPS",
    Params_catlog %in% names(Raf_OA_phenotype_data_outlier_filtered)
    [which(names(Raf_OA_phenotype_data_outlier_filtered) == "Knee_pain")] ~ "Knee pain",
    TRUE ~ "Other"  
  ))

unique(Raf_OA_phenotype_data_long$Params_catlog)

write.csv(Raf_OA_phenotype_data_long , 'Raf_OA_phenotype_data_long.csv')

### Box Plots for each phenotype ###
Raf_phenotype_data_cytokines_figure <- Raf_OA_phenotype_data_long  %>%
  filter(Params_type == "Serum Proinflammatory Markers") %>%
  ggplot(aes(x = Months, y = Params_Value, fill = Group)) +
 stat_boxplot(geom ='errorbar') + 
  geom_boxplot(alpha = 1, outlier.shape = NA) +
  geom_point(position = position_jitterdodge(jitter.width = 0.25, dodge.width = 0.75), 
             size = 2, color = "black", shape = 21) +
  scale_fill_manual(values = c("grey", "#FFA07A", "dimgrey", "tomato", "black", "red")) +
  mytheme3 +
  facet_wrap(~Params_catlog, scales = "free", ncol = 3) +  
  ylab("Serum cytokines (pg/mL)") +
  xlab("Time (month)")
Raf_phenotype_data_cytokines_figure

Raf_phenotype_data_SCFA_figure <- Raf_OA_phenotype_data_long  %>%
  filter(Params_type == "SCFA" & Params_catlog %in% c("Acetate", "Propionate", "Butyrate")) %>%
  ggplot(aes(x = Months, y = Params_Value, fill = Group)) +
 stat_boxplot(geom ='errorbar') + 
  geom_boxplot(alpha = 1, outlier.shape = NA) +
  geom_point(position = position_jitterdodge(jitter.width = 0.25, dodge.width = 0.75), 
             size = 2, color = "black", shape = 21) +
  scale_fill_manual(values = c("grey", "#FFA07A", "dimgrey", "tomato", "black", "red")) +
  mytheme3 +
  facet_wrap(~Params_catlog, scales = "free", ncol = 3) +  
  ylab("Fecal SCFA (umol/g)") +
  xlab("Time (month)")
Raf_phenotype_data_SCFA_figure

Raf_OA_LPS_conc_plot <- ggplot(Raf_OA_phenotype_data_outlier_filtered, 
                                   aes(x = Months, y = LPS, fill = Group)) +
  stat_boxplot(geom ='errorbar') + 
  geom_boxplot(alpha = 1, outlier.shape = NA) +
  geom_point(position = position_jitterdodge(jitter.width = 0.25, dodge.width = 0.75), 
             size = 2, color = "black", shape = 21) +
  scale_fill_manual(values = c("grey", "#FFA07A", "dimgrey", "tomato", "black", "red")) +
  mytheme3 +
  ylab("Concentration (umol/g)") +
  xlab("Time (month)") 

Raf_OA_LPS_conc_plot

ggsave("Raf_OA_LPS_conc_plot.svg", 
       plot = Raf_OA_LPS_conc_plot, 
       device = "svg", 
       width = 2, height = 4, 
       units = "in")

ggsave("Raf_phenotype_data_SCFA.svg", 
       plot = Raf_phenotype_data_SCFA_figure, 
       device = "svg", 
       width = 4.5, height = 3, 
       units = "in")

ggsave("Raf_phenotype_cytokines_raw.svg", 
       plot = Raf_phenotype_data_cytokines_figure, 
       device = "svg", 
       width = 5, height = 5, 
       units = "in")

